---
title: "Analysis of Restaurant Tips Dataset"
author: "Ozan Tivriz"
date: "2025-11-01"
output: html_document
---

This is an analysis of a restaurant and its customers, aimed at understanding 
the relationship between tips, group sizes, and days.

## Importing Data

First, we load the libraries that we will use. Tidyverse is
general analysis package, conflicted helps us escape conflictions and
scales for readable graphic scales. After that we import the data.
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
knitr::opts_knit$set(root.dir = "C:/Users/Ozan Tivriz/Desktop/data")
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale("LC_TIME", "English_United States") 
```

```{r importing data, message=FALSE, warning=FALSE}
library(tidyverse)
tips <- read.csv("restaurant_tips/tip.csv")
```

## Cleaning Data

When we look at the day column in the data, there is only 4 inputs.
```{r}
table(tips$day)
```

We assume that the restaurant is only open for four days a week and rename the 
days for a cleaner look. A new data frame is created to avoid affecting the
original dataset.
```{r new df, message=FALSE, warning=FALSE}
tips_v2 <- tips %>% 
  mutate(day = recode(day, 
                       "Fri" = "Friday",
                       "Sat" = "Saturday",
                       "Sun" = "Sunday",
                       "Thur" = "Thursday"))
```

We also create a vector to order the days and attach this factor to our data.
```{r factor new df, message=FALSE, warning=FALSE}
days <- c("Thursday","Friday","Saturday","Sunday")
tips_v2$day <- factor(tips_v2$day, levels = days)
```

There is also a sex column and a size column, which indicates how many people
are dining together.
To analyze tipping behavior by gender, we first need to create a new data frame
containing only solo diners.

However, there are only four people who dined alone, so any analysis based on 
this data would be biased. Therefore, we decided not to include gender-specific
analysis.
```{r}
table(tips_v2$size)
```

## Analyze

A new data frame is created to analyze the total bill. Many similar data frames
will be created throughout this analysis, but for readability, we will not show
every step to keep the report concise.
```{r total bill med mean df, message=FALSE, warning=FALSE}
mean_total_bill <- tips_v2 %>%
  group_by(size) %>%
  summarise(
    mean_total_bill = mean(total_bill),
    median_total_bill = median(total_bill)
  )
```

The bar chart illustrates how the average and median total bill change 
depending on the group size.
```{r total bill med mean chart, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(mean_total_bill, aes(x=size, y=mean_total_bill)) +
  geom_bar(stat = "identity",fill = "#86CD82", alpha = 0.7) +
  geom_point(aes(y=median_total_bill), size = 2, color = "red") +
  geom_text(aes(y=median_total_bill, label = round(median_total_bill,1)),
            vjust = -0.9, color = "red") +
  labs(title = "Average and Median Total Bill by Group Size",
       x = "Group Size",
       y = "Total Bill") +
  scale_x_continuous(breaks = 1:6) + 
  theme_minimal()

```

```{r tips med mean df, message=FALSE,echo=FALSE,warning=FALSE}
mean_tips <- tips_v2 %>% 
  group_by(size) %>% 
  summarise(
    mean_tip=mean(tip),
    median_tip=median(tip)
  )

```

The bar chart illustrates how the average and median tips change 
depending on the group size.
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplot(mean_tips, aes(x= size, y= mean_tip)) +
  geom_bar(stat = "identity",fill="#86CD82", alpha = 0.7) +
  geom_point(aes(y=median_tip), size = 2, color ="red") +
  geom_text(aes(y=median_tip, label = round(median_tip,1)),
            vjust = -0.9, color = "red") +
  labs(title = "Average and Median Tips by Group Size",
       x = "Group Size",
       y= "Tip Amount",
       fill = "Group Size") +
  scale_x_continuous(breaks = 1:6) +
  theme_minimal()
```

We also calculate and visualize the data based on the tip rate.
```{r tip rate df, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
tips_v2 <- tips_v2 %>% 
  mutate(tip_rate = tip / total_bill)

tip_rate <- tips_v2 %>% 
  group_by(size) %>% 
  summarise(
    mean_tip_rate = mean(tip_rate),
    median_tip_rate = median(tip_rate)
  )

```
```{r tip rate bar-chart, message=FALSE,echo=FALSE,warning=FALSE}
ggplot(tip_rate, aes(x= size, y = mean_tip_rate)) +
  geom_bar(stat = "identity",fill = "#86CD82", alpha = 0.7) +
  geom_point(aes(y = median_tip_rate), size = 2, color = "red") +
  geom_text(aes(y = median_tip_rate, label = round(median_tip_rate, 3)),
            vjust = -1.6, color = "red") +
  labs(title = "Average and Median Tip Rate by Group Size",
       x = "Group Size",
       y= "Tip Rate",
       fill = "Group Size") +
  scale_x_continuous(breaks = 1:6) +
  theme_minimal()
```

We first visualize the average per-person bill for each group size.
```{r per person df and med mean, warning=FALSE,echo=FALSE,message=FALSE}
tips_v2 <- tips_v2 %>% 
  mutate(per_person_bill = total_bill/size)

mean_person_bill <- tips_v2 %>% 
  group_by(size) %>% 
  summarise(mean_per_person = mean(per_person_bill))
```
```{r per person line graph, echo=FALSE,message=FALSE,warning=FALSE}
ggplot(mean_person_bill) +
  geom_line(aes(x=size, y=mean_per_person), size = 0.7) +
  labs(title = "Per Person Bill / Group Size",
       x = "Group Size",
       y= "Average Per Person Bill") +
  scale_x_continuous(breaks = 1:6) +
  ylim(0,NA) +
  theme_minimal()
```

Next, we examine the individual per-person bills, showing their distribution
and a linear trend line to capture the overall relationship.
```{r per person trend line graph, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(tips_v2, aes(x= size, y= per_person_bill)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous(breaks = 1:6) +
  ylim(0,NA) +
  labs(title = "Per Person Bill vs Group Size",
       x = "Group Size",
       y = "Per Person Bill")+
  theme_minimal()
```

The correlation between group size and per-person bill is -0.175, which
indicates a weak negative relationship. This suggests that, on average, as
the group size increases, the amount each person pays slightly decreases.
```{r correlation per person, echo=FALSE, warning=FALSE}
cor(tips_v2$size, tips_v2$per_person_bill)
```

The correlation between group size and tip amount is 0.489, indicating a
moderate positive relationship. This suggests that larger groups tend to
leave higher tips in absolute terms.
```{r correlation tip, echo=FALSE,warning=FALSE}
cor(tips_v2$size, tips_v2$tip)
```

With a correlation of -0.143, the tip rate slightly decreases as group size 
increases, showing a very weak negative relationship.
```{r correlation tip rate, echo=FALSE, warning=FALSE}
cor(tips_v2$size, tips_v2$tip_rate)
```

To check for differences between days, we calculate the average and median tips for each day.
```{r tips by day df, echo=FALSE,warning=FALSE,message=FALSE}

mean_tips_day <- tips_v2 %>% 
  group_by(day) %>% 
  summarise(
    mean_tip=mean(tip),
    median_tip=median(tip)
  )

```
```{r tips by day bar-chart, echo=FALSE,message=FALSE,warning=FALSE}
ggplot(mean_tips_day, aes(x= day, y= mean_tip, fill=day)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_point(aes(y=median_tip), size = 2, color ="red") +
  geom_text(aes(y=median_tip, label = round(median_tip,1)),
            vjust = -0.9, color = "red") +
  labs(title = "Average and Median Tips by Day",
       x = "Day of Week",
       y= "Tip Amount",
       fill = "Days") +
  theme_minimal()
```

Looking at this graph, we can see that customer groups of more than two people
are more likely to visit the restaurant on weekends.
```{r group sizes compared by day, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(tips_v2, aes(x = size, fill = day)) +
  geom_bar() +
  facet_wrap(~day,  strip.position = "bottom") +
  theme_minimal() +
  theme(
    strip.placement = "outside",
    axis.text.x = element_text(size = 10),
    strip.background = element_blank()
  ) +
  labs(
    title = "Group Sizes Compared by Day",
    x = "Group Size",
    y = "Count"
  )
```

Here we compare customers by smokers and non-smokers.
```{r smoker df, echo=FALSE,warning=FALSE,message=FALSE}
tip_rate_smoker <- tips_v2 %>% 
  group_by(smoker) %>% 
  summarise(
    mean_tip_rate = mean(tip_rate),
    median_tip_rate = median(tip_rate)
  )
```
```{r smoker med mean, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(tip_rate_smoker, aes(x=smoker , y = mean_tip_rate)) +
  geom_bar(stat = "identity",fill = "#86CD82", alpha = 0.7) +
  geom_point(aes(y = median_tip_rate), size = 2, color = "red") +
  geom_text(aes(y = median_tip_rate, label = round(median_tip_rate, 3)),
            vjust = -0.6, color = "red") +
  labs(title = "Average and Median Tip Rate by Smokers",
       x = "Smoker",
       y= "Tip Rate",
       fill = "Smoker") +
  ylim(0,0.2)+
  theme_minimal()
```

Smokers compared by day.

```{r smokers compared by day, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(tips_v2, aes(x = smoker, fill = day)) +
  geom_bar() +
  facet_wrap(~day,  strip.position = "bottom") +
  theme_minimal() +
  theme(
    strip.placement = "outside",
    axis.text.x = element_text(size = 10),
    strip.background = element_blank()
  ) +
  labs(
    title = "Smokers Compared by Day",
    x = "Smokers",
    y = "Count"
  )

```

## Conclusion

\* No strong correlations were observed between group size and total bill 
per person.

\* As the group size increases, the total tip amount tends to increase; however,
the tip rate slightly decreases, indicating that larger groups leave higher tips
in absolute terms but slightly lower percentages of their total bill.

\* Groups larger than two people are more likely to visit the restaurant on 
weekends.

\* Additionally, tip rates do not appear to be related to customers’ smoking
status.